Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation

### 2024-05-10 mop/draft.tex
Updates from Overleaf

\begin{lem}[Change of Particle Measure]
    \label{lem:change-measure-proper-weights}
    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$. Now, let $\{(Y_j^J,V_j^J),j=1,\dots,J\}$ be a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is represented, on average, $\pi^J_j J$ times. Write
    \[
    (Y_j^J,V_j^J) = \big(\tilde X^J_{a(j)},U^J_{a(j)}/\pi^J_{a(j)}\big),
    \]
    where $a$ is called the ancestor function. If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j),j=1,\dots,J\}$ targets $f_X$.
\end{lem}

@@

By a similar decomposition to that of the error of the $k$-truncated gradient to the MOP-$\alpha$ gradient estimate,
\begin{align*}
    &\left\lVert\nabla_\theta\hat\ell^1(\theta) - \nabla_\theta\hat\ell^\alpha(\theta) \right\rVert_\infty\\
    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_\infty\\
    &= \Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\\
    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_\infty \\
    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)
    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha}\right)\Bigg)\right\rVert_\infty \\
    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)\right|
    +\left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha}\right)\right|\Bigg)\right\rVert_\infty.
\end{align*}

@@

@@

We decompose
$$\E||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty \leq \E||\nabla_\theta\ell_\theta - \nabla_\theta\hat s_k^1(\theta)||_\infty + \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty + \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty.$$

The broad idea is that the first term can be controlled by a mixing argument, the second term garners $O(Nk^2)$ error, and the third term is $O\left(\frac{N}{1-\alpha}\right)$. It is unclear whether the bias can be reduced further -- the gradient is a sum of $N$ terms, and is therefore $O(N)$ itself.

The first term -- bound ground truth of conditional scores from particle approximation conditional on correct filtering distribution at time $n-k$, bound that from particle approximation conditional on arbitrary filtering distribution at time $n-k$. First term sums up $k$ terms, $O(Nk^2)$ error, second term bounded by $k$ applications of the Markov kernel, so just apply forgetting of the particle filter result.

But you can get an MSE bound like this! Simply decompose
$$||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 \leq ||\nabla_\theta\ell_\theta - \nabla_\theta\hat s_k^1(\theta)||_2^2 + ||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2 + ||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_2^2,$$
and observe that 
$$||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty =  \left\lVert\sum_{n=1}^N \nabla_\theta \hat\ell_n^\alpha(\theta) - \sum_{n=1}^N \nabla_\theta \ell_n(\theta)\right\rVert_\infty \leq \sum_{n=1}^N\left\lVert \nabla_\theta \hat\ell_n^\alpha(\theta) -  \nabla_\theta \ell_n(\theta)\right\rVert_\infty.$$

>>>>

\begin{lem}[Change of Particle Measure]
    \label{lem:change-measure-proper-weights}
    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$. Now, let $\{(Y_j^J,V_j^J),j=1,\dots,J\}$ be a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is represented, on average, proportional to $\pi^J_j J$ times. Write
    \[
    (Y_j^J,V_j^J) = \big(\tilde X^J_{a(j)},U^J_{a(j)}/\pi^J_{a(j)}\big),
    \]
    where $a$ is called the ancestor function. If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j),j=1,\dots,J\}$ targets $f_X$.
\end{lem}

@@

By a similar decomposition to that of the error of the $k$-truncated gradient to the MOP-$\alpha$ gradient estimate,
\begin{align*}
    &\left\lVert\nabla_\theta\hat\ell^1(\theta) - \nabla_\theta\hat\ell^\alpha(\theta) \right\rVert_\infty\\
    &= \left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_\infty\\
    &= \Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)- \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\\
    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=1}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=1}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_\infty \\
    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)
    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha}\right)\Bigg)\right\rVert_\infty \\
    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)\right|
    +\left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha}\right)\right|\Bigg)\right\rVert_\infty.
\end{align*}

@@

\kevin{This subsection is not useless I think! It shows that the bias vanishes continuously as $\alpha$ goes to $1$.}

@@

We decompose the bias $||\nabla_\theta\ell(\theta) - \E\nabla_\theta \hat\ell^\alpha(\theta)||_\infty \leq \E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty$ into
$$\E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty \leq \E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_\infty + \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty + \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty.$$

The broad idea is that the first term can be controlled by a mixing argument, the second term garners $O(Nk)$ error, and the third term is $O\left(\frac{N}{1-\alpha}\right)$. It is unclear whether the bias can be reduced further -- the gradient is a sum of $N$ terms, and is therefore $O(N)$ itself.

The first term is the problem term. We will first bound the ground truth of conditional scores from particle approximation conditional on correct filtering distribution at time $n-k$, and then bound that from the particle approximation resulting from an arbitrary filtering distribution at time $n-k$. That is, where $\hat\pi_n$ is the particle approximation at time $n$ of the posterior $\pi_n$ and we write $\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)$ for the truncated MOP-$1$ conditional score estimate given the correct filtering distribution at time $n-k$,

\begin{align*}
    &\E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_\infty \\
    &\leq \sum_{n=1}^N \E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_\infty \\
    &\leq \sum_{n=1}^N \E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_\infty + \sum_{n=1}^N \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_\infty.
\end{align*}

The first term is a particle approximation dependent on $k$ timesteps, so by Lemma 2 of \cite{karjalainen2023}, this is bounded by
$$\E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_\infty \leq \frac{CG'(\theta)k}{\sqrt{J}}.$$

The second term amounts to bounding the difference between functionals of two different particle measures that mix under the same Markov kernel. Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing. We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator, 
$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
i.e. the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.

Then, we can bound 
$\E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_\infty$ by
$$\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}} = \beta_{TV}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
implying that
\begin{align*}
    \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_\infty 
    &\lesssim \sup_{\mu, \nu} \sup_{||\psi||_\infty \leq 1/2} G'(\theta)\left|(\mu \textbf{M}_{n,n+k})(\psi)-(\nu \textbf{M}_{n,n+k})(\psi)\right| \\
    &\leq \sup_{\mu, \nu} G'(\theta)\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\|_{\mathrm{TV}} \\
    &\leq G'(\theta)(1-\epsilon)^{\floor{k/(c\log(J))}} ||\hat\pi_{n-k} - \pi_{n-k}||_{\text{TV}} \\
    &\leq G'(\theta)(1-\epsilon)^{\floor{k/(c\log(J))}} \sup_{||\psi||_{\infty} \leq 1/2} |\hat\pi_{n-k}(\psi) - \pi_{n-k}(\psi)| \\
    &\lesssim G'(\theta)(1-\epsilon)^{\floor{k/(c\log(J))}}.
\end{align*}

Therefore, we have that
\begin{align*}
    &\E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_\infty \\
    &\leq \sum_{n=1}^N \E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_\infty + \sum_{n=1}^N \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_\infty\\
    &\lesssim N\frac{CG'(\theta)k}{\sqrt{J}} + NG'(\theta)(1-\epsilon)^{\floor{k/(c\log(J))}}.
\end{align*}

Now that the first term $\E||\nabla_\theta\ell(\theta)$ is taken care of, it remains to bound $$\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty \text{ and } \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty.$$

We see, by a similar argument to the variance bound, that
\begin{align*}
    &\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty\\
    &= \E\left\lVert\sum_{n=1}^N \nabla_\theta \log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)}\right) - \sum_{n=1}^N \nabla_\theta\log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right) \right\rVert_\infty\\
    &= \E\Bigg\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)- \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\\
    &\qquad\qquad -\log\left(\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right) + \log\left(\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}\right)\Bigg)\Bigg\rVert_\infty \\
    &= \E\left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1,k}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)
    -\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1,k}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\Bigg)\right\rVert_\infty \\
    &\leq \sum_{n=1}^N \E\left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1,k}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right|
    +\left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1,k}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\right|\Bigg)\right\rVert_\infty.
\end{align*}

Each of these terms can be bounded using the log-derivative trick and that $w_{n,j}^{F,\theta,1,k} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
\begin{align*}
    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,1,k}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_{\infty}\\
    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,1,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,1,k}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}}}\right\rVert_{\infty} \\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,1,k}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,\alpha,k}\right\rVert_{\infty} \\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,1,k}}{w_{n,j}^{F,\theta,1,k}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,\alpha,k}}{w_{n,j}^{F,\theta,\alpha,k}}\right\rVert_{\infty} \\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,1,k}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_{\infty} \\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta,1,k}\right)-\log\left(w_{n,j}^{F,\theta,\alpha,k}\right)\right)\right\rVert_{\infty}\\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=n-k}^n\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) - \sum_{i=n-k}^n{\alpha^{(n-i)}}\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)\right)\right\rVert_{\infty} \\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\sum_{i=n-k}^{n}(1-\alpha^{(n-i)})\log\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right) \right)\right\rVert_{\infty}\\
    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})\left\lVert\nabla_\theta\log\left(g_{i,j}^{A,F,\theta} \right)\right\rVert_{\infty}\\
    &\leq \frac{1}{J}\sum_{j=1}^J \sum_{i=n-k}^{n}(1-\alpha^{(n-i)})G'(\theta)\\
    &\leq G'(\theta)\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right),
\end{align*}
where the second-last line follows from Assumption \ref{assump:bounded-measurement}.

So, 
\begin{align*}
    &\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty\\
    &\leq \sum_{n=1}^N \E\left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1,k}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right|
    +\left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1,k}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\right|\Bigg)\right\rVert_\infty \\
    &\leq 2NG'(\theta)\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right) 
    \leq 2NG'(\theta)k.
\end{align*}

Now, the third term, $\E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty$, fortunately has already been dealt with. From the variance argument, we know that
$$||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty \leq 2\frac{\alpha^k}{1-\alpha}NG'(\theta),$$
and taking expectations on both sides yields our desired result.

Therefore, our decomposition yields the bias bound
\begin{align*}
    &||\nabla_\theta\ell(\theta) - \E\nabla_\theta \hat\ell^\alpha(\theta)||_\infty \leq \E||\nabla_\theta\ell(\theta) - \nabla_\theta \hat\ell^\alpha(\theta)||_\infty \\
    &\leq \E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_\infty + \E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_\infty + \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_\infty \\
    &\lesssim \min_{k \leq N} \left(N\frac{CG'(\theta)k}{\sqrt{J}} + NG'(\theta)(1-\epsilon)^{\floor{k/(c\log(J))}} + 2NG'(\theta)\left(k-\frac{\alpha(1-\alpha^k)}{1-\alpha}\right) + 2\frac{\alpha^k}{1-\alpha}NG'(\theta)\right) \\
    &\lesssim \min_{k \leq N} NG'(\theta)\left(\frac{k}{\sqrt{J}}+(1-\epsilon)^{\floor{k/(c\log(J))}}+k+\frac{\alpha^k  + \alpha^{k+1} - \alpha}{1-\alpha}\right).
\end{align*}

\subsection{Bounding the MSE of MOP-$\alpha$}

\kevin{First term sums up $k$ terms, $O(Nk^2)$ error, second term bounded by $k$ applications of the Markov kernel, so just apply forgetting of the particle filter result. }

We can obtain an $O(N)$ MSE bound from an argument that is similar to the bias argument. Simply observe that 
$$||\nabla_\theta\ell_\theta - \nabla_\theta \hat\ell^\alpha(\theta)||_2^2 =  \left\lVert\sum_{n=1}^N \nabla_\theta \hat\ell_n^\alpha(\theta) - \sum_{n=1}^N \nabla_\theta \ell_n(\theta)\right\rVert_2^2 \leq \sum_{n=1}^N\left\lVert \nabla_\theta \hat\ell_n^\alpha(\theta) -  \nabla_\theta \ell_n(\theta)\right\rVert_2^2,$$

and decompose
$$\sum_{n=1}^N\left\lVert \nabla_\theta \hat\ell_n^\alpha(\theta) -  \nabla_\theta \ell_n(\theta)\right\rVert_2^2 \leq \sum_{n=1}^N||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2 + \sum_{n=1}^N||\nabla_\theta\hat s_{n,k}^1(\theta) - \nabla_\theta\hat s_{n,k}^\alpha(\theta)||_2^2 + \sum_{n=1}^N||\nabla_\theta\hat s_{n,k}^\alpha(\theta) -  \nabla_\theta\hat\ell_n^\alpha(\theta)||_2^2.$$

As before, we bound the first term, $\sum_{n=1}^N||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2$, by decomposing it into two terms, 
$$\sum_{n=1}^N||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2 \leq \sum_{n=1}^N ||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_2^2 + \sum_{n=1}^N ||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2.$$

The first term is a particle approximation dependent on $k$ timesteps, so by Lemma 2 of \cite{karjalainen2023}, this is bounded by
$$\E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_2^2 \leq \frac{CpG'(\theta)^2k^2}{J}.$$

The second term amounts to bounding the difference between functionals of two different particle measures that mix under the same Markov kernel. Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing. We know from Theorem 3 of \cite{karjalainen2023} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator, 
$$\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
i.e. the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.

Then, we can bound 
$\E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2$ by
$$\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}} = \beta_{TV}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},$$
implying that
\begin{align*}
    \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2 
    &\lesssim \sup_{\mu, \nu} \sup_{||\psi||_\infty \leq 1/2} pG'(\theta)^2\left|(\mu \textbf{M}_{n,n+k})(\psi)-(\nu \textbf{M}_{n,n+k})(\psi)\right| \\
    &\leq \sup_{\mu, \nu} pG'(\theta)^2\|\mu \textbf{M}_{n,n+k}-\nu \textbf{M}_{n,n+k}\|_{\mathrm{TV}} \\
    &\leq pG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}} ||\hat\pi_{n-k} - \pi_{n-k}||_{\text{TV}} \\
    &\leq pG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}} \sup_{||\psi||_{\infty} \leq 1/2} |\hat\pi_{n-k}(\psi) - \pi_{n-k}(\psi)| \\
    &\lesssim pG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}}.
\end{align*}

Therefore, we have that
\begin{align*}
    &\E||\nabla_\theta\ell(\theta) - \nabla_\theta\hat s_k^1(\theta)||_2^2 \\
    &\leq \sum_{n=1}^N \E||\nabla_\theta\ell_n(\theta) - \nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta)||_2^2 + \sum_{n=1}^N \E||\nabla_\theta\hat s_{n,k|\hat\pi_{n-k}=\pi_{n-k}}^1(\theta) - \nabla_\theta\hat s_{n,k}^1(\theta)||_2^2\\
    &\lesssim N\frac{CpG'(\theta)^2k^2}{J} + NpG'(\theta)^2(1-\epsilon)^{\floor{k/(c\log(J))}}.
\end{align*}

Now that the first term $\E||\nabla_\theta\ell(\theta)||_2^2$ is taken care of, it remains to bound $$\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2 \text{ and } \E||\nabla_\theta\hat s_k^\alpha(\theta) -  \nabla_\theta\hat\ell^\alpha(\theta)||_2^2.$$

We see, by a similar argument to the bias bound, that
\begin{align*}
    &\E||\nabla_\theta\hat s_k^1(\theta) - \nabla_\theta\hat s_k^\alpha(\theta)||_2^2\\
    &\leq \sum_{n=1}^N \E\left\lVert\nabla_\theta \Bigg(\left|\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,1,k}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right|
    +\left|\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,1,k}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\right|\Bigg)\right\rVert_2^2.
\end{align*}
